Supplemental Online Materials:\\\textit{Double Standards in Judging Collective Action}

### 2023-04-19 Documents/manuscript.Rmd
switch to Quarto

@@

Item response theory is a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. <!-- As such, it allows us to disentangle how acceptable protest actions are from how accepting respondents are of them and, in so doing, to detect double standards in judging collective action. --> When respondents judge whether each of a set of actions is an acceptable means of protest (1 = *yes*, 0 = *no*), a two-parameter logistic item response theory model estimates responses as a function of several latent (unobserved) parameters: 
$$ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\Pr ( y_{ij} = 1 )$ is the probability that respondent $j$ considers action $i$ an acceptable means of protest, $\theta_j$ estimates how *accepting* respondent $j$ is of various protest actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

>>>>

### Are Partisan Double Standards Based in Ideology or Identity?

So far, we ...

<!-- Throughout this paper, we will use "identity-based" as a shorthand for double standards based on social categories such as race, gender, and class and "ideology-based" as a shorthand for double standards based on system-justification and political ideology. -->

@@

Item response theory is a conceptual and statistical framework for understanding how the characteristics of both items and respondents shape responses to a set of items [@demars_irt_2010]. <!-- As such, it allows us to disentangle how acceptable protest actions are from how accepting respondents are of them and, in so doing, to detect double standards in judging collective action. --> When respondents judge whether each of a set of actions is an acceptable means of protest (1 = *yes*, 0 = *no*), a two-parameter logistic item response theory model estimates responses as a function of several latent (unobserved) parameters: 
$$ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) $$
where $\Pr ( y_\textit{ij} = 1 )$ is the probability that respondent $j$ considers action $i$ an acceptable means of protest, $\theta_j$ estimates how *accepting* respondent $j$ is of various protest actions, $\beta_i$ estimates how *acceptable* action $i$ is, and $\alpha_i$ estimates how *discriminating* action $i$ is.

### 2023-04-19 Documents/manuscript.tex
switch to Quarto

% Font
\usepackage[utf8]{inputenc}
\usepackage{MinionPro} 
\input{glyphtounicode}
\pdfgentounicode=1 
\usepackage{microtype}

% Format
\usepackage[letterpaper, margin = 1in]{geometry}
\setlength{\columnsep}{18pt}
\setcounter{secnumdepth}{0}
\usepackage[compact]{titlesec}
\titleformat*{\section}{\centering\normalfont\Large\bfseries}
\titleformat*{\subsection}{\raggedright\normalfont\normalsize\bfseries}
\titleformat*{\subsubsection}{\raggedright\normalfont\normalsize\bfseries\itshape}
\usepackage{authblk}
\renewcommand\Affilfont{\small}
\newcommand{\refsection}{\section{References}}
\providecommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Links
\usepackage[colorlinks = true, linkcolor = black, urlcolor = black, citecolor = black]{hyperref}

@@

@@

% Frontmatter
\title{ Double Standards in Judging Collective Action\thanks{} }
\author[1]{Nils K. Reimer}
\affil[1]{University of California, Santa Barbara, United States}
\author[2]{Marija Brankovi\'{c}}
\affil[2]{Singidunum University and University of Belgrade, Serbia}
\author[3]{Iniobong Essien}
\affil[3]{Leuphana University of L\"{u}neburg, Germany}
\author[4]{Jin X. Goh}
\affil[4]{Colby College, United States}
\author[5]{S\'{e}bastien Goudeau}
\affil[5]{University of Poitiers, France}
\author[6]{N\'{o}ra A. Lantos}
\affil[6]{E\"{o}tv\"{o}s Lor\'{o}nd University, Hungary}
\author[7]{Jenny Veldman}
\affil[7]{NYU Abu Dhabi, United Arab Emirates}

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In three preregistered experiments, we then use this instrument
to test for double standards in judging protest actions for workers'
rights in the United Kingdom and for or against defunding the police and
for or against restricting abortion in the United States. We find
evidence for ideology-based, but not for identity-based, double
standards: People judge the same protest actions to be more acceptable
when the protesters' cause aligns with their own ideological
positions---but do not show consistent ingroup bias when judging
collective action by ingroup and outgroup members.

@@

Collective action---that is, any action taken by group members to
advance a shared political goal (for similar definitions, see Becker,
2012; van Zomeren, 2016)---is a powerful force driving social change.
For example, Black Lives Matter protests lastingly shifted public
discourse about racial inequity toward antiracist ideas (Dunivin et al.,
2022).

@@

In this research, we move this distinction from the researcher's
intuition into the realm of scientific investigation and test for double
standards in where people draw the line between acceptable (normative)
and unacceptable (non-normative) forms of collective action. For that
purpose, we define a double standard as judging the same action as more
or less acceptable depending on who is performing the action and for
what reason (for an analogous definition, see Foschi, 2000).\footnote{Feinberg
  et al. (2020) provided incidental evidence for double standards in
  judging collective action as, in a manipulation check, Black
  participants rated Black Lives Matter protests as less extreme than
  White participants and liberal participants rated protest actions for
  progressive causes as less extreme than conservative participants.} In
this section, we derive predictions from social psychological theories
about how identity-based and ideology-based double standards could lead
people to judge the same protest actions as more or less acceptable
depending on who the protesters are and what they are protesting.

@@

In the same vain, we propose that ingroup bias results in identity-based
double standards in judging collective action by ingroup and outgroup
members. That is, we hypothesize that people will judge the same protest
actions as more acceptable if the protesters are ingroup members
(\emph{Hypothesis 1a}) or if the cause of the protest aligns with the
ingroup's interests (\emph{Hypothesis 1b}). This hypothesis could
explain why, for example, far more Black than White Americans considered
it appropriate for Black athletes to kneel down to protest police
violence against Black Americans (YouGov, 2017). Notably, this
hypothesis applies equally to historically disadvantaged groups
mobilizing against social injustice and to historically advantaged
groups defending their group's position.

@@

As system-justifying beliefs are closely related to political
conservatism (Jost et al., 2003), this hypothesis could explain partisan
differences in what actions liberals and conservatives consider
acceptable means of protest for progressive and conservative causes.
Ideology-based double standards could, for example, explain why far more
Democrats than Republicans considered it appropriate to kneel during the
national anthem to protest racist police violence (YouGov, 2017).

@@

Item response theory is a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). When respondents judge
whether each of a set of actions is an acceptable means of protest (1 =
\emph{yes}, 0 = \emph{no}), a two-parameter logistic item response
theory model estimates responses as a function of several latent
(unobserved) parameters:
\[ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\Pr ( y_{ij} = 1 )\) is the probability that respondent \(j\)
considers action \(i\) an acceptable means of protest, \(\theta_j\)
estimates how \emph{accepting} respondent \(j\) is of various protest
actions, \(\beta_i\) estimates how \emph{acceptable} action \(i\) is,
and \(\alpha_i\) estimates how \emph{discriminating} action \(i\) is.

@@

We included additional measures to describe the sample, to describe
reactions to the manipulation, or to use in non-preregistered analyses.
In addition to demographic questions, we asked participants how outraged
they were about recent incidents of police violence against Black
Americans, to what extent they identifed with the protesters described
in the study, and to what extent they identifed with their racial
ingroup (1 = \emph{not at all}, 7 = \emph{very much}). We also asked how
often, if at all, participants had participated in protest actions such
as the ones we had asked about (1 = \emph{never}, 5 = \emph{very often})
and whether they has participated in protests for reforming, defunding
or abolishing the police; against reforming, defunding or abolishing the
police; or in neither. We measured political orientation with a one-item
semantic differential scale: ``People often describe their political
orientation as liberal or conservative. On a scale from liberal to
conservative, where would you position yourself?'' (1 = \emph{liberal},
7 = \emph{conservative}).

@@

We preregistered the sample size, hypotheses, inclusion/exclusion
criteria, measures, and manipulations
(\url{https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822}).
We deviated from the preregistration in two ways. First, we stopped
collecting data before reaching the preregistered sample size. We did so
as the Supreme Court decision on reproductive rights, announced on June
24, 2022, rendered the experimental manipulation invalid. Second, we do
not report results for two further hypotheses---testing moral conviction
as a potential mechanism underlying ideology-based double standards---as
those findings are beyond the scope of this manuscript and will be
reported in a future publication. We made all materials, data, and
analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standards when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggest that, unlike judgments in other domains, judgments about
collective action are not subject to double standards based on group
identities such as class, race, or gender. Why did we not find any
evidence for identity-based double standards?

@@

Our findings provided a relevant comparison to findings from a recent
experiment in which White Americans viewed pictures of either peaceful
or destructive Black Lives Matter protests and rated how legitimate
(i.e., fair, reasonable, legitimate, and justified) they considered the
pictured protests to be (Teixeira et al., 2022). In this experiment,
more conservative participants rated destructive and, to a lesser
extent, peaceful protests to be less legitimate than more liberal
participants. Likewise, participants who identified more strongly with
their ethnic ingroup rated destructive and, to a lesser extent, peaceful
protests to be less legitimate than participants who identified less
strongly with their ingroup. However, participants' group identification
was no longer associated with their protest ratings after controlling
for their political orientation.\footnote{These findings are based on
  our reanalysis, reported in the SOM, of Teixeira et al.'s (2022)
  experiment. We thank the authors for making their data available
  online.} In this way, results from the experiment aligned with our
conclusion that political ideologies are more important than social
identities for shaping how people judge controversial forms of
collective action.

@@

>>>>

% Font
\usepackage[MnSymbol]{mathspec}
\setmainfont{Minion Pro}
\setmathsfont(Digits,Latin,Greek)
[Numbers={Lining,Proportional}]{Minion Pro}
\usepackage{microtype}

% Format
\usepackage[letterpaper, margin = 1in]{geometry}
\setlength{\columnsep}{18pt}
\setcounter{secnumdepth}{0}
\usepackage[compact]{titlesec}
\titleformat*{\section}{\centering\normalfont\Large\bfseries}
\titleformat*{\subsection}{\raggedright\normalfont\normalsize\bfseries}
\titleformat*{\subsubsection}{\raggedright\normalfont\normalsize\bfseries\itshape}
\usepackage{authblk}
\renewcommand\Affilfont{\small}

% References/Footnotes
\newcommand{\refsection}{\section{References}}

% Header
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{14.5pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{ DOUBLE STANDARDS }
\fancyhead[R]{\thepage}

% Links
\usepackage[colorlinks = true, linkcolor = black, urlcolor = black, citecolor = black]{hyperref}
\usepackage{xurl}

@@

% Lists
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

% Front matter
\title{ Double Standards in Judging Collective Action\thanks{} }
\author[1]{Nils K. Reimer}
\affil[1]{University of California, Santa Barbara, United States}
\author[2]{Marija Brankovi\'{c}}
\affil[2]{Singidunum University and University of Belgrade, Serbia}
\author[3]{Iniobong Essien}
\affil[3]{Leuphana University of L\"{u}neburg, Germany}
\author[4]{Jin X. Goh}
\affil[4]{Colby College, United States}
\author[5]{S\'{e}bastien Goudeau}
\affil[5]{University of Poitiers, France}
\author[6]{N\'{o}ra A. Lantos}
\affil[6]{E\"{o}tv\"{o}s Lor\'{o}nd University, Hungary}
\author[7]{Jenny Veldman}
\affil[7]{NYU Abu Dhabi, United Arab Emirates}

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In three preregistered experiments, we then use this instrument
to test for double standards in judging protest actions for workers'
rights in the United Kingdom and for or against defunding the police and
for or against restricting abortion in the United States. We find
evidence for ideology-based, but not for identity-based, double
standards: People judge the same protest actions to be more acceptable
when the protesters' cause aligns with their own ideological
positions---but do not show consistent ingroup bias (in terms of class,
race, and gender) when judging collective action by ingroup and outgroup
members.

@@

Collective action---that is, any action taken by group members to
advance a shared political goal (for similar definitions, see Becker,
2012; van Zomeren, 2016)---is a powerful force driving social change.
For example, Black Lives Matter protests shifted public discourse about
racial inequity toward antiracist ideas (Dunivin et al., 2022).

@@

In this research, we move this distinction from the researcher's
intuition into the realm of scientific investigation and test for double
standards in where people draw the line between acceptable (normative)
and unacceptable (non-normative) forms of collective action. For that
purpose, we define a double standard as judging the same protest actions
as more or less acceptable depending on who the protesters are and what
they are protesting. (for an analogous definition, see Foschi,
2000).\footnote{Feinberg et al. (2020) provided incidental evidence for
  double standards in judging collective action as, in a manipulation
  check, Black participants rated Black Lives Matter protests as less
  extreme than White participants and liberal participants rated protest
  actions for progressive causes as less extreme than conservative
  participants.}

@@

In the same vein, we propose that ingroup bias results in identity-based
double standards in judging collective action by ingroup and outgroup
members. That is, we hypothesize that people will judge the same protest
actions as more acceptable if the protesters are ingroup members
(\emph{Hypothesis 1a}) or if the cause of the protest aligns with the
ingroup's interests (\emph{Hypothesis 1b}). This hypothesis could
explain why, for example, far more Black than White Americans considered
it appropriate for Black athletes to kneel down to protest police
violence against Black Americans (YouGov, 2017). Notably, this
hypothesis applies equally to historically disadvantaged groups
mobilizing against social injustice and to historically advantaged
groups defending their group's position.

@@

As system-justifying beliefs are closely related to political
conservatism (Jost et al., 2003), this hypothesis could explain partisan
differences in what actions liberals and conservatives consider
acceptable means of protest for progressive and conservative
causes---for example, why far more Democrats than Republicans considered
it appropriate to kneel during the national anthem to protest racist
police violence (YouGov, 2017).

\hypertarget{are-partisan-double-standards-based-in-ideology-or-identity}{%
\subsubsection{Are Partisan Double Standards Based in Ideology or
Identity?}\label{are-partisan-double-standards-based-in-ideology-or-identity}}

\begin{itemize}
\tightlist
\item
  Are Partisan Double Standards Based in Ideology or Identity?
\item
  Evidence for partisanship as a elective affinity (Jost, 2021)
\item
  Evidence for partisanship as a social identity
\item
  We will use ``identity-based'' as a shorthand for double standards
  based on assigned categories such as class, race, and gender and
  ``ideology-based'' as a shorthand for double standards based on
  system-justification and political orientation.
\end{itemize}

@@

Item response theory is a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). When respondents judge
whether each of a set of actions is an acceptable means of protest (1 =
\emph{yes}, 0 = \emph{no}), a two-parameter logistic item response
theory model estimates responses as a function of several latent
(unobserved) parameters:
\[ \Pr ( y_\textit{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\Pr ( y_\textit{ij} = 1 )\) is the probability that respondent
\(j\) considers action \(i\) an acceptable means of protest,
\(\theta_j\) estimates how \emph{accepting} respondent \(j\) is of
various protest actions, \(\beta_i\) estimates how \emph{acceptable}
action \(i\) is, and \(\alpha_i\) estimates how \emph{discriminating}
action \(i\) is.

@@

We included additional measures to describe the sample, to describe
reactions to the manipulation, or to use in non-preregistered analyses.
In addition to demographic questions, we asked participants how outraged
they were about recent incidents of police violence against Black
Americans, to what extent they identifed with the protesters described
in the study, and to what extent they identifed with their racial
ingroup (1 = \emph{not at all}, 7 = \emph{very much}). We also asked how
often, if at all, participants had participated in protest actions such
as the ones we had asked about (1 = \emph{never}, 5 = \emph{very often})
and whether they had participated in protests for reforming, defunding
or abolishing the police; against reforming, defunding or abolishing the
police; or in neither. We measured political orientation with a one-item
semantic differential scale: ``People often describe their political
orientation as liberal or conservative. On a scale from liberal to
conservative, where would you position yourself?'' (1 = \emph{liberal},
7 = \emph{conservative}).

@@

We preregistered the sample size, hypotheses, inclusion/exclusion
criteria, measures, and manipulations
(\url{https://osf.io/kphyw/?view_only=59f5792a6d3f4fa48516ea9d5f637822}).
We deviated from the preregistration in two ways. First, we stopped
collecting data before reaching the preregistered sample size. We did so
as the Supreme Court decision on reproductive rights, announced on June
24, 2022, rendered the experimental manipulation invalid. Second, we do
not report results for two further hypotheses---testing moral conviction
as a potential mechanism underlying ideology-based double standards---as
those findings are beyond the scope of this manuscript and will be
reported in a future publication. We made all materials, data, and
analysis scripts available online
(\url{https://osf.io/d3yev/?view_only=40782034017c40f0bcecb1cc87760b62}).
Our sample size was determined by budget and time constraints.

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standards when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggest that, unlike judgments in other domains, judgments about
collective action are not subject to double standards based on group
identities such as class, race, or gender.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-jost_left_2021}{}}%
Jost, J. T. (2021). \emph{Left and right: The psychological significance
of a political distinction}. Oxford University Press.

### 2023-04-19 Documents/supplemental-materials.tex
switch to Quarto

% Font
\usepackage[utf8]{inputenc}
\usepackage{MinionPro} 
\input{glyphtounicode}
\pdfgentounicode=1 
\usepackage{microtype}

@@

>>>>

% Font
\usepackage[MnSymbol]{mathspec}
\setmainfont{Minion Pro}
\setmathsfont(Digits,Latin,Greek)
[Numbers={Lining,Proportional}]{Minion Pro}
\usepackage{microtype}

@@

\setcounter{table}{3}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 1}
\centering
\begin{tabular}{rlrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 \\ \midrule
 1 & Proportion            &      & -.31 & -.16 & -.23 \\
 2 & Political Orientation & -.31 &      &  .41 &  .41 \\
 3 & Social Dominance      & -.16 &  .41 &      &  .29 \\
 4 & System Justification  & -.23 &  .41 &  .29 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Political Orientation (1 = \textit{left}, 7 = \textit{right}); Social Dominance Orientation (\textit{factor score}); Systen-Justifying Beliefs (\textit{factor score})}
\end{table}

\setcounter{table}{4}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 2}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.21 &  .29 & -.23 & -.07 \\
 2 & System Justification  & -.21 &      & -.45 &  .47 &  .04 \\
 3 & Policy Support        &  .29 & -.45 &      & -.49 &  .00 \\
 4 & Political Orientation & -.23 &  .47 & -.49 &      &  .03 \\
 5 & Identification        & -.07 &  .04 &  .00 &  .03 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Racial Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}

\setcounter{table}{5}

\begin{table}[t!]
\caption{Correlations between variables in Experiment 3}
\centering
\begin{tabular}{rlrrrrr}
\toprule
\# & Variable              &    1 &    2 &    3 &    4 &    5 \\ \midrule
 1 & Proportion            &      & -.27 & -.22 & -.32 & -.11 \\
 2 & System Justification  & -.27 &      &  .61 &  .78 &  .16 \\
 3 & Policy Support        & -.22 &  .61 &      &  .74 &  .16 \\
 4 & Political Orientation & -.32 &  .78 &  .74 &      &  .20 \\
 5 & Identification        & -.11 &  .16 &  .16 &  .20 &      \\ \bottomrule
\addlinespace
\end{tabular}
\caption*{\textit{Note.} Proportion of actions considered acceptable (0--1); Systen-Justifying Beliefs (\textit{factor score}); Policy Support (1 = \textit{strongly oppose}, 5 = \textit{strongly support}); Political Orientation (1 = \textit{liberal}, 7 = \textit{conservative}); Gender Identification (1 = \textit{not at all}, 7 = \textit{very much})}
\end{table}
